---
title: "Soil data and how it varies across sites"
author: "Scott Klasek"
date: "2022-12-19"
output: github_document
---
#### purpose
To see how different soil data varies across states and develop hypotheses about how it could be driving trends in microbiomes across sites.  

#### load libraries
```{r}
library(tidyverse)
# library(soiltexture)
library(ggtern)
library(patchwork)
```

#### load and combine data
```{r}
soil <- read.csv(file = "/Users/klas0061/Desktop/UMN/jim_info/soil_characteristics_jim_2023_06_13.csv", na.strings = c(".")) # convert . to NA
table(soil$State) # no USDA ME2 site

larkin.data <- read.csv(file = "/Users/klas0061/Desktop/UMN/larkin_info/chem_data_first_two_years.csv") # ME2 data that I could find (this csv was manually curated)

soil <- bind_rows(soil, larkin.data) # bind in ME2 data with the rest of the site data
```

#### plotting edaphic factors that vary noticeably across sites
Note: I tried to write these ggplots as functions, using embracing to specify each variable to plot. But it didn't work, because REASONS. 
```{r}
cols.alphabetical.east.west <- c("#DC4405", "#B3A369", "#236192", "#7A0019", "#FFC72A", "#C5050C", "#18453B", "#B0D7FF", "gray") # color scale

# plotting pH in samples from the first two years
ggph <- ggplot(soil %>% filter(Year < 21), aes(State, pH, color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+ggtitle("A")+theme(legend.position = "none")

# plotting organic matter and C in samples from the first two years
ggom <- ggplot(soil %>% filter(Year < 21), aes(State, OM...., color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+scale_y_continuous("% Organic matter")+ggtitle("B")+theme(legend.position = "none")

ggtc <- ggplot(soil %>% filter(Year < 21), aes(State, Total.C...., color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+scale_y_continuous("% total C")+ggtitle("C")+theme(legend.position = "none")

# ggom + ggtc + plot_layout(guides = "collect") 

# NPK
ggn <- ggplot(soil %>% filter(Year < 21), aes(State, Nitrate.N..ppm., color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+scale_y_continuous("Nitrate (ppm)")+ggtitle("D")+theme(legend.position = "none")

ggp <- ggplot(soil %>% filter(Year < 21), aes(State, P.Bray..ppm., color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+scale_y_continuous("P (ppm, Bray)")+ggtitle("E")+theme(legend.position = "none")

ggk <- ggplot(soil %>% filter(Year < 21), aes(State, K..ppm., color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+scale_y_continuous("K (ppm)")+ggtitle("F")+theme(legend.position = "none")

# CEC 
ggcec <- ggplot(soil %>% filter(Year < 21), aes(State, CEC..meq.100g., color = factor(State, levels = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))))+
  geom_jitter()+
  scale_color_manual("Site", values = cols.alphabetical.east.west)+
  scale_x_discrete("", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"))+
  theme_bw()+scale_y_continuous("Cation Exchange Capacity \n(mEq/100g)")+ggtitle("G")+theme(legend.position = "none")


ggph + ggom + ggtc + ggn + ggp + ggk + ggcec
```

CO and ID clearly have higher pH than the rest, ME has highest OM.  

Other noteworthy trends:

ME is a bit of an outlier, with highest OM and total C, also high nitrate and P. 

CO has high K and nitrate.  

OR has high K.  

MN and ND (both MN really) all have way higher total C than WI does. Microbial communities from these sites all cluster together.  


#### Make a ternary plot of soil texture classifications   
Help from [this resource here](https://saryace.github.io/flipbook_soiltexture_en/#37).  
```{r}
soil.texture <- soil %>% select(State, Sand.., Silt.., Clay..) %>% drop_na() # trim dataframe and remove NAs 
colnames(soil.texture)[2:4] <- c("Sand", "Silt", "Clay") # clean up column names

theme_set(theme_bw()) # call a theme somehow
data(USDA) # load USDA soil classification polygon boundaries
USDA_text <- USDA  %>% group_by(Label) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE) # 

# plot with USDA classifications
tplot <- ggplot(data = USDA, aes(y = Clay, x = Sand, z = Silt)) +
  coord_tern(L = "x", T = "y", R = "z") +
  geom_polygon(
    aes(fill = Label), alpha = 0.0, size = 0.5,
    color = "black") +
  geom_text(data = USDA_text, aes(label = Label), color = 'black', size = 2) +
  geom_point(data = soil.texture, aes(x = Sand, y = Clay, z = Silt, color = State, alpha = 0.2))+
  scale_color_manual("Site", limits = c("OR", "ID", "CO", "MN", "ND", "WI", "MI", "ME", "US"), labels = c("OR", "ID", "CO", "MN1", "MN2", "WI", "MI", "ME1", "ME2"), values = cols.alphabetical.east.west)+guides(fill="none", alpha = "none")
tplot
```

MI has a lot of textural variability. It doesn't seem to impact fungal beta-diversity, but I wonder if this explains the high bacterial beta-diversity. 


#### session info
```{r}
sessionInfo()
```