---
title: "table info"
author: "Scott Klasek"
date: "2023-01-03"
output: github_document
---
#### Purpose
To make pretty tables. JK, the first table will show numbers of microbiome samples from each site x year x season x amplicon, and the second will show cultivar info by site.   

#### libraries
```{r}
packages <- c("tidyverse", "phyloseq", "plyr")
t <- lapply(packages, require, character.only = TRUE) # load all packages at once
```

#### load data
```{r}
all.fung.1920.c.ps <- readRDS(file = "/Users/klas0061/Desktop/UMN/phyloseqs/fung1920.cleaned.ps") # cleaned fungal phyloseq object
core.bact.ps <- readRDS(file = "/Users/klas0061/Desktop/UMN/phyloseqs/core/core.bact.asvs.ps") # core bacterial phyloseq object to save memory
```

#### not a table anymore, but counts of ITS and 16S communities by site.  
```{r}
its.site <- table(data.frame(all.fung.1920.c.ps@sam_data)$state) # ITS microbiome counts by state
bact.site <- table(data.frame(core.bact.ps@sam_data)$state) # 16S microbiome counts by state
site.counts <- rbind("ITS count" = its.site, "16S count" = bact.site) # bind its and 16s counts
colnames(site.counts) <- c("CO", "ID", "ME1", "MI", "MN1", "MN2", "OR", "ME2", "WI") # correct site labels
site.counts <- site.counts[,c(7,2,1,5,6,9,4,3,8)] # order west to east 
site.counts
```


#### table S1
Counts of ITS and 16S communities by site, year, and season.  
```{r}
its.counts <- ddply(data.frame(all.fung.1920.c.ps@sam_data), .(year, season, state),nrow) # sum up ITS counts by year, season, and state
colnames(its.counts)[4] <- "ITS count"

bact.counts <- ddply(data.frame(core.bact.ps@sam_data), .(year, season, state),nrow) # sum up 16S counts by year, season, and state
colnames(bact.counts)[4] <- "16S count"

# merge them together and make the table neat
both.counts <- left_join(bact.counts, its.counts, by = c("year", "season", "state")) %>% drop_na(state) # remove na states (blanks and tech reps)
both.counts$year <- paste(20, both.counts$year, sep = "") # add 20 to year
both.counts <- both.counts %>% arrange(state, year, factor(season, levels = c("Spring", "Summer", "Fall"))) # order by season
both.counts[,2] <- ifelse(both.counts[,2] == "Spring", "Pre-plant",
                          ifelse(both.counts[,2] == "Summer", "60 days after planting", 
                                 ifelse(both.counts[,2] == "Fall", "After harvest", NA))) # change season labels to harvest descriptions
both.counts[which(both.counts[,3] == "ME"),3] <- "ME1" # update site labels
both.counts[which(both.counts[,3] == "US"),3] <- "ME2"
both.counts[which(both.counts[,3] == "MN"),3] <- "MN1"
both.counts[which(both.counts[,3] == "ND"),3] <- "MN2"
colnames(both.counts)[3] <- "site"
both.counts
```

We are missing: ID, MI, and WI fall 2019. And we have an extra MN fall (2018). 

#### table S2
Counts of 16S communities from each potato cultivar at each site.  
```{r}
cultivar.count <- ddply(data.frame(core.bact.ps@sam_data), .(cultivar, state),nrow) # get counts of cultivars
colnames(cultivar.count)[c(2,3)] <- c("site","count") # fix colnames

cultivar.count[which(cultivar.count[,2] == "ME"),2] <- "ME1" # update site labels
cultivar.count[which(cultivar.count[,2] == "US"),2] <- "ME2"
cultivar.count[which(cultivar.count[,2] == "MN"),2] <- "MN1"
cultivar.count[which(cultivar.count[,2] == "ND"),2] <- "MN2"

cultivar.count <- pivot_wider(cultivar.count, names_from = "cultivar", values_from = "count") # pivot wider, cultivars as columns
cultivar.count <- cultivar.count[1:9,1:10] # cut out NAs
cultivar.count <- data.frame(t(column_to_rownames(cultivar.count, "site"))) # put site as row name, transpose, and convert to df
cultivar.count <- cultivar.count[,c(7,4,3,1,2,8,6,5,9)] # place sites in west-east order
cultivar.count[is.na(cultivar.count)] <- 0 # turn NAs into zeros
cultivar.count
```
All sites used Burbank. ME2 samples were between rows of different varieties. Norkotah is popular in the west, Bannock in MN, and Caribou in ME. All other varieties were grown in one site only. 

#### session info
```{r}
sessionInfo()
```